---
title: "Text prediction - milestone assignment of the capstone project"
author: "Evgeniy Zabrodskiy"
date: "17 March 2016"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis
The goal of this analysis is to understand the distribution and relationship between the words, tokens, and phrases in the text in order to build a predictive model.
The distributions of frequencies of words, word pairs and word triplets are analised and shown in form of histograms, word clouds and barplots.

```{r, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(tm)
library(RWeka)
library(wordcloud)
library(ggplot2)
library(grid)
library(gridExtra)
library(stringi)

# wordcloud/RWeka does not work with parallel processing when NGramTokenizer is used.
options(mc.cores=1)
set.seed(4444)
```

## Loading data
```{r Loading, cache = TRUE, echo = FALSE}
# this is needed for tokenizer in DocumentTermMatrix to work. For some reason does not with parallel processing.
# Found solution here: http://stackoverflow.com/questions/17703553/bigrams-instead-of-single-words-in-termdocument-matrix-using-r-and-rweka

ds <- DirSource(directory = "final/en_US/")
docs <- Corpus(ds, readerControl = list(reader = readPlain, language = "en", load = TRUE))

docs.num <- length(docs)
docs.name <- c("blogs", "news", "twitter")

# gather simple information about documents corpus
docs.info.df <- data.frame(FileName = character(), 
                           NumberOfRows = integer(),
                           NumberOfWords = integer())
for(i in 1:docs.num) {
    docs.info.df <- rbind(docs.info.df, 
                          data.frame(FileName = docs[[i]]$meta$id, 
                                     NumberOfRows = length(docs[[i]]$content),
                                     NumberOfWords = sum(stri_count_words(docs[[i]]$content))))
}
```

Here are the loaded file names, their lenghts in number of rows and total number of words:
```{r, echo = FALSE}
docs.info.df
```

## Sampling and cleaning data
The files are quite big (more than 4 millions of rows in total) and building document term matrix using original files is computationally expensive and may take a lot of time. Besides, for the word prediction task we're interested in frequent unigrams, bigrams and trigrams which can be obtained from a smaller sample.  

```{r getSample, echo = FALSE}
sample.size <- as.integer(10000)
# copy original doc for processing (don't know how to make it more efficient yet)
docs.sample <- docs
# select random sample of sample.size from the original
# iterate through documents in the corpus
for (i in 1:docs.num) {
    # get length of each document (number of lines)
    l <- length(docs[[i]]$content)
    # get random sample of lines in the document and replace the original doc in the corpus with the smaller sample.
    docs.sample[[i]]$content <- docs[[i]]$content[sample(1:l, as.integer(sample.size))]
}
```

We randomly selected **`r sample.size`** rows from each document and all further cleaning and transformations are done using the sample.  

```{r cleanDocs, echo = FALSE}
# remove punctuation
docs.sample <- tm_map(docs.sample, removePunctuation)   
# remove numbers
docs.sample <- tm_map(docs.sample, removeNumbers)
# remove unnecessary whitespaces
docs.sample <- tm_map(docs.sample, stripWhitespace)
# to lowecase
docs.sample <- tm_map(docs.sample, content_transformer(tolower))
# finalize preprocessing
docs.sample <- tm_map(docs.sample, content_transformer(PlainTextDocument))
```

Cleaning procedures and transformations that were applied to the sample are:  
- remove punctuation,  
- remove numbers,  
- convert to lowercase,  
- remove unnecessary whitespaces.  

```{r DocumentTermMatrix, echo = FALSE}
##  create document term matrix for all documents
bigramTokenizer <- function(x) NGramTokenizer(x, 
                                Weka_control(min = 2, max = 2))
trigramTokenizer <- function(x) NGramTokenizer(x, 
                                Weka_control(min = 3, max = 3))

dtm.sample.l <- list(unigram = DocumentTermMatrix(docs.sample),
                     bigram = DocumentTermMatrix(docs.sample, control = list(tokenize = bigramTokenizer)),
                     trigram = DocumentTermMatrix(docs.sample, control = list(tokenize = trigramTokenizer)))

# remove sparse terms
dtms.sample.l <- lapply(dtm.sample.l, function(x) removeSparseTerms(x, 0.1))
```

## Exploratory analysis

### Basic information about the data
Once the data is clean, the document term matrix can be created. It contains the frequncies of terms in each document. The frequencies of terms can be converted to probabilities for prediction modelling at later stages of the project.  

Here is the default output of the document term matrix created from the data sample:  

```{r, echo = FALSE}
# gather information about sample corpus
docs.sample.info.df <- data.frame(FileName = character(), 
                                  NumberOfRows = integer(),
                                  NumberOfWords = integer(),
                                  NumberOfTermsPerDocument = integer())
for(i in 1:docs.num) {
    docs.sample.info.df <- rbind(docs.sample.info.df, 
                                 data.frame(FileName = docs.sample[[i]]$meta$id, 
                                            NumberOfRows = length(docs.sample[[i]]$content),
                                            NumberOfWords = sum(dtm.sample.l$unigram$v[dtm.sample.l$unigram$i == i]),
                                            NumberOfTermsPerDocument = sum(dtm.sample.l$unigram$v != 0 & 
                                                                               dtm.sample.l$unigram$i == i)))
}

# show output of DocumentTermMatrix with sparse terms
dtm.sample.l$unigram
```

From the output we can see some useful information, including total number of terms.  
Here are the lengths of samples, number of words instances, number of terms (unique words) per each document:  

```{r, echo = FALSE}
# show calculated info
docs.sample.info.df
```

### Questions of the analysis
#### 1. Some words are more frequent than others - what are the distributions of word frequencies?  

```{r Freq1, echo = FALSE}
# Create matrixes
dtms.sample.m.l <- list(unigram = as.matrix(dtms.sample.l$unigram),
                        bigram = as.matrix(dtms.sample.l$bigram),
                        trigram = as.matrix(dtms.sample.l$trigram))

# sort frequencies of terms in separate documents
# input - document term matrix
getTopTerms <- function(ngram.m, n = 20) {

    # initialize structures for top list
    df <- data.frame(Term = factor(), Freq = integer(), Doc = factor())
    top <- df

    for(i in 1:docs.num) {
        # Unigrams
        doc <- data.frame(Term = dimnames(ngram.m)$Terms, 
                          Freq = ngram.m[i, ], 
                          Doc = factor(docs.name[i], levels = docs.name))
        df <- rbind(df, doc[order(doc$Freq, decreasing = TRUE), ])
    
        top <- rbind(top, df[as.integer(df$Doc) == i, ][1:n, ])
    }
    # return result
    top
}

# get top for all n-grams
ngram.df.top.l <- lapply(dtms.sample.m.l, function(x) getTopTerms(x, 20))

# iterate over (unigram, bigram, trigram) separate dataframe by doc id
ngram.df.top.l.docs <- lapply(ngram.df.top.l, function(x) split(x, x$Doc))

# function that reorders levels according to the order of terms corresponding 
# to ordered frequencies
orderLevels <- function(ngram.docs.l) { 
    lapply(ngram.docs.l, function(y) {
            y$Term <- factor(y$Term, 
                             levels = unique(as.character(y$Term)))
            y
        })
}

# iterate over (unigrams, bigrams, trigrams) and apply orderLevels which iterates 
# over docs.
ngram.df.top.l.docs <- lapply(ngram.df.top.l.docs, orderLevels)

# calculate aggregated frequency
freq.l <- lapply(dtms.sample.m.l, colSums)
# sort by most frequently used words
freq.l <- lapply(freq.l, function(x) sort(x, decreasing = TRUE))
terms.l <- lapply(freq.l, names)

# dataframe for more convenient use in ggplot
freq.df.l <- list(unigram = data.frame(Term = factor(terms.l$unigram, 
                                                     levels = terms.l$unigram), 
                                       Freq = freq.l$unigram),
                  bigram = data.frame(Term = factor(terms.l$bigram, 
                                                    levels = terms.l$bigram), 
                                      Freq = freq.l$bigram),
                  trigram = data.frame(Term = factor(terms.l$trigram, 
                                                     levels = terms.l$trigram), 
                                       Freq = freq.l$trigram))
```

```{r withSparseFreq}
# using dtm with sparse terms
dtm.sample.unigram.m <- as.matrix(dtm.sample.l$unigram)
# calculate frequency
frequency <- colSums(dtm.sample.unigram.m)
# sort by most frequently used words
frequency <- sort(frequency, decreasing = TRUE)
terms <- names(frequency)

sparse.freq.df <- data.frame(Term = factor(terms, levels = terms), 
                             Freq = frequency)
```

Here are the histograms to understand the distribution of word frequencies. Due to the distribution properties, it does not look good without transformations and in order to get a better looking histograms, the distribution is shown on the log scale.  
The histogram on the left shows the distribition of word frequencies including sparse terms, the histogram on the right shows the same kind of distribution but with sparse terms removed from the matrix.  
We can see that the more frequent the word is (to the right on the histogram), the less number of words with such a high frequency there are.  

```{r displayHist1, echo = FALSE, fig.width = 9, fig.height = 3}
hist.sparse <- qplot(log(Freq), 
                     data = sparse.freq.df, 
                     geom = "histogram", 
                     bins = 20, 
                     fill=..count..) +
    labs(x = "log(Term Frequency)",
         y = "Count",
         title = "Frequencies of words")

# terms frequencies histogram
hist.wosparse <- qplot(log(Freq), 
                       data = freq.df.l$unigram, 
                       geom = "histogram", 
                       bins = 20, 
                       fill=..count..) +
    labs(x = "log(Term Frequency)",
         y = "Count",
         title = "Frequencies without sparse terms")

hists <- arrangeGrob(hist.sparse, hist.wosparse, ncol = 2)
grid.draw(hists)
```

Below are the barplots showing unigram terms frequencies per each document (Blogs, News and Twitter) separately and aggregated for all three documents (lower-right graph).  

```{r displayFreq1, fig.width = 9, fig.height = 7, echo = FALSE}
# graphs with term frequencies per document
g.perdoc1.bar <- ggplot(data = ngram.df.top.l.docs$unigram$blogs, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Blogs") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc2.bar <- ggplot(data = ngram.df.top.l.docs$unigram$news, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in News") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc3.bar <- ggplot(data = ngram.df.top.l.docs$unigram$twitter, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Twitter") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

# graph with most frequent terms aggregated for all document
g.agg.bar <- ggplot(data = freq.df.l$unigram[1:20, ], 
                    aes(Term, Freq, fill = Freq)) +
    geom_bar(stat = "identity") +
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in all documents") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

graphs <- arrangeGrob(g.perdoc1.bar, g.perdoc2.bar, g.perdoc3.bar, g.agg.bar, ncol = 2, nrow = 2)
grid.draw(graphs)
```

For quick perception of the terms frequencies one can have a look at words cloud:  

```{r displayWC1, echo = FALSE}
# draw wordcloud
wordcloud(words = terms.l$unigram[1:100], 
          freq = freq.l$unigram[1:100], 
          random.order = FALSE, 
          scale = c(8, 1),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE,
          colors = brewer.pal(8, "Dark2"))
```

#### 2. What are the frequencies of 2-grams and 3-grams in the dataset?  
Similar to unigram terms, here are the barplots for bigrams:  

```{r displayFreq2, fig.width = 9, fig.height = 7, echo = FALSE}
## Bigrams
# graphs with term frequencies per document
g.perdoc1.bar <- ggplot(data = ngram.df.top.l.docs$bigram$blogs, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Blogs") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc2.bar <- ggplot(data = ngram.df.top.l.docs$bigram$news, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in News") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc3.bar <- ggplot(data = ngram.df.top.l.docs$bigram$twitter, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Twitter") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

# graph with most frequent terms aggregated for all document
g.agg.bar <- ggplot(data = freq.df.l$bigram[1:20, ], 
                    aes(Term, Freq, fill = Freq)) +
    geom_bar(stat = "identity") +
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in all documents") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

graphs <- arrangeGrob(g.perdoc1.bar, g.perdoc2.bar, g.perdoc3.bar, g.agg.bar, ncol = 2, nrow = 2)
grid.draw(graphs)
```

```{r displayWC2, echo = FALSE, eval = FALSE}
# draw wordcloud
wordcloud(words = terms.l$bigram[1:100], 
          freq = freq.l$bigram[1:100], 
          random.order = FALSE, 
          scale = c(4, 0.3),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE,
          colors = brewer.pal(8, "Dark2"))
```

Same way of presentation for trigram term frequencies. The difference of most frequent terms between different documents (types of sources) is significant especially copared to unigram frequencies above.  

```{r displayFreq3, fig.width = 9, fig.height = 7, echo = FALSE}
## Trigrams
# graphs with term frequencies per document
g.perdoc1.bar <- ggplot(data = ngram.df.top.l.docs$trigram$blogs, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Blogs") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc2.bar <- ggplot(data = ngram.df.top.l.docs$trigram$news, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in News") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

g.perdoc3.bar <- ggplot(data = ngram.df.top.l.docs$trigram$twitter, 
                       aes(Term, Freq, fill = Freq)) + 
    geom_bar(stat = "identity") + 
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in Twitter") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

# graph with most frequent terms aggregated for all document
g.agg.bar <- ggplot(data = freq.df.l$trigram[1:20, ], 
                    aes(Term, Freq, fill = Freq)) +
    geom_bar(stat = "identity") +
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms in all documents") +
    theme(axis.text.x=element_text(angle = 75, hjust = 1))

graphs <- arrangeGrob(g.perdoc1.bar, g.perdoc2.bar, g.perdoc3.bar, g.agg.bar, ncol = 2, nrow = 2)
grid.draw(graphs)
```

```{r displayWC3, echo = FALSE, eval = FALSE}
# draw wordcloud
wordcloud(words = terms.l$trigram[1:100], 
          freq = freq.l$trigram[1:100], 
          random.order = FALSE, 
          scale = c(3, 0.2),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE,
          colors = brewer.pal(8, "Dark2"))
```

#### 3. How many unique words do you need in a frequency sorted dictionary to cover 50% of all word instances in the language? 90%?  
```{r, echo = FALSE}
# reusing data calculated in withSparseFreq section
words.total <- sum(frequency)
words.agg <- 0
words.count <- 1

while(words.agg < 0.5 * words.total) {
    words.agg <- words.agg + frequency[words.count]
    words.count <- words.count + 1
}

words.count50 <- words.count

while(words.agg < 0.9 * words.total) {
    words.agg <- words.agg + frequency[words.count]
    words.count <- words.count + 1
}

words.count90 <- words.count
```

Number of frequent words covering half of all word instances in the language: **`r as.integer(words.count50)`**  
Number of frequent words covering 90% of all word instances in the language: **`r as.integer(words.count90)`**  

#### 4. How do you evaluate how many of the words come from foreign languages?  

One of the ways to identify foreign words is by looking up for a word in a language dictionary. This approach is quite straighforward and has disadvantages such as incorrect classification of misspelled words.  
Another possible way is using machine learning algorithms with language profiles. Similar approach is used in *langid* library.  

#### 5. Can you think of a way to increase the coverage -- identifying words that may not be in the corpora or using a smaller number of words in the dictionary to cover the same number of phrases?  

Identifying words that may not be in the corpora can be done by an algorithm which uses external dictionary with linguistic markers such as type of the word (noun, verb, adjective, adverb, etc.) together with some machine learning algorithms that can suggest the word.  
Using a smaller number of words in the dictionary to cover the same number of phrases can be possible by stemming words and suggesting endings based on some algorithm which takes into account grammar rules of the language.  

## References
1. [Text Mining Infrastructure in R](https://www.jstatsoft.org/article/view/v025i05)
2. [Basic Text Mining in R](https://rstudio-pubs-static.s3.amazonaws.com/31867_8236987cf0a8444e962ccd2aec46d9c3.html)
3. [Natural Language Processing](https://www.coursera.org/course/nlp)
