---
title: "Text prediction - milestone assignment of the capstone project"
author: "Evgeniy Zabrodskiy"
date: "8 March 2016"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis
The goal of this analysis is to understand the distribution and relationship between the words, tokens, and phrases in the text in order to build a predictive model.
The distributions of frequencies of words, word pairs and word triplets are analised and shown in form of word clouds and barplots.

## Initialization

```{r}
library(tm)
library(RWeka)
library(wordcloud)
library(ggplot2)
library(grid)
library(gridExtra)

# wordcloud/RWeka does not work with parallel processing when NGramTokenizer is used.
options(mc.cores=1)
```

### Loading data
```{r Loading, cache = TRUE}
# this is needed for tokenizer in DocumentTermMatrix to work. For some reason does not with parallel processing.
# Found solution here: http://stackoverflow.com/questions/17703553/bigrams-instead-of-single-words-in-termdocument-matrix-using-r-and-rweka

ds <- DirSource(directory = "final/en_US/")
docs <- Corpus(ds, readerControl = list(reader = readPlain, language = "en", load = TRUE))
summary(docs)
```

### Sampling and cleaning data
```{r getSample}
sample.size <- 10000
# copy original doc for processing (don't know how to make it more efficient yet)
docs.sample <- docs
# select random sample of sample.size from the original
# iterate through documents in the corpus
for (i in 1:length(docs)) {
    # get length of each document (number of lines)
    l <- length(docs[[i]]$content)
    # get random sample of lines in the document and replace the original doc in the corpus with the smaller sample.
    docs.sample[[i]]$content <- docs[[i]]$content[sample(1:l, as.integer(sample.size))]
}
```

```{r cleanDocs}
# remove punctuation
docs.sample <- tm_map(docs.sample, removePunctuation)   

# remove numbers
docs.sample <- tm_map(docs.sample, removeNumbers)

# remove some special characters
for(j in seq(docs.sample)) {
    docs.sample[[j]]$content <- gsub("/", " ", docs.sample[[j]]$content)   
    docs.sample[[j]]$content <- gsub("@", " ", docs.sample[[j]]$content)   
    docs.sample[[j]]$content <- gsub("\\|", " ", docs.sample[[j]]$content)
}   

# remove stopwords
# docs.sample <- tm_map(docs.sample, removeWords, stopwords("english"))

# to lowercase
docs.sample <- tm_map(docs.sample, tolower)  

# remove unnecessary whitespaces
docs.sample <- tm_map(docs.sample, stripWhitespace)

# finalize preprocessing
docs.sample <- tm_map(docs.sample, PlainTextDocument)
```

```{r DocumentTermMatrix}
##  create document term matrix for all documents
# Singles
dtm.sample.unigram <- DocumentTermMatrix(docs.sample)

# remove sparse terms
dtms.sample.unigram <- removeSparseTerms(dtm.sample.unigram, 0.1)

# Doubles
bigramTokenizer <- function(x) NGramTokenizer(x, 
                                Weka_control(min = 2, max = 2))

dtm.sample.bigram <- DocumentTermMatrix(docs.sample, control = list(tokenize = bigramTokenizer))

# remove sparse terms
dtms.sample.bigram <- removeSparseTerms(dtm.sample.bigram, 0.1)

# Triples
trigramTokenizer <- function(x) NGramTokenizer(x, 
                                Weka_control(min = 3, max = 3))

dtm.sample.trigram <- DocumentTermMatrix(docs.sample, control = list(tokenize = trigramTokenizer))

# remove sparse terms
dtms.sample.trigram <- removeSparseTerms(dtm.sample.trigram, 0.1)
```

## Exploratory analysis

1. Some words are more frequent than others - what are the distributions of word frequencies?  

```{r displayFreq1}

## Single words
# create matrix
dtms.sample.unigram.m <- as.matrix(dtms.sample.unigram)

# sort frequencies of terms in separate documents
dtms.sample.unigram.df <- data.frame(Term = factor(), Freq = integer(), Doc = factor())
dtms.sample.unigram.df.top <- dtms.sample.unigram.df
for(i in 1:nrow(dtms.sample.unigram.m)) {
    doc <- data.frame(Term = dimnames(dtms.sample.unigram.m)$Terms, 
                      Freq = dtms.sample.unigram.m[i, ], 
                      Doc = as.factor(i))
    dtms.sample.unigram.df <- rbind(dtms.sample.unigram.df, 
                                    doc[order(doc$Freq, decreasing = TRUE), ])

    dtms.sample.unigram.df.top <- rbind(dtms.sample.unigram.df.top, 
                                        dtms.sample.unigram.df[dtms.sample.unigram.df$Doc == i, ][1:20, ])
}

dtms.sample.unigram.df.top <- dtms.sample.unigram.df.top[order(dtms.sample.unigram.df.top$Freq, decreasing = TRUE), ]
# drop unused levels and reorder according to ordered frequencies
dtms.sample.unigram.df.top$Term <- factor(dtms.sample.unigram.df.top$Term, 
                                          levels = unique(as.character(dtms.sample.unigram.df.top$Term)))

# graph with term frequencies per document
g.perdoc.bar <- ggplot(data = dtms.sample.unigram.df.top, 
                       aes(Term, Freq, fill = Doc)) + 
    geom_bar(stat = "identity") + 
    facet_grid(Doc ~ .) +
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms per documents") +
    theme(axis.text.x=element_text(angle = 60, hjust = 1))

# calculate aggregated frequency
frequency1 <- colSums(dtms.sample.unigram.m)

# sort by most frequently used words
frequency1 <- sort(frequency1, decreasing = TRUE)
words1 <- names(frequency1)

# dataframe for more convenient use in ggplot
frequency1.df <- data.frame(w = factor(words1, 
                                      levels = words1), 
                            f = frequency1)

# graph with most frequent terms aggregated for all document
g.agg.bar <- ggplot(data = frequency1.df[1:20, ], 
                    aes(w, f, fill = f)) +
    geom_bar(stat = "identity") +
    labs(x = "Terms",
         y = "Frequency",
         title = "Most frequent terms among all documents") +
    theme(axis.text.x=element_text(angle = 60, hjust = 1))

graphs <- arrangeGrob(g.perdoc.bar, g.agg.bar, nrow = 2)
grid.draw(graphs)

# draw wordcloud
g.agg.wc <- wordcloud(words = words1[1:100], 
          freq = frequency1[1:100], 
          random.order = FALSE, 
          scale = c(8, 1),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE,
          colors = brewer.pal(8, "Dark2"))
```

2. What are the frequencies of 2-grams and 3-grams in the dataset?  

```{r diplayFreq2}
## Bigrams

dtms.sample.bigram.m <- as.matrix(dtms.sample.bigram)

# calculate frequency
frequency2 <- colSums(dtms.sample.bigram.m)

# sort by most frequently used words
frequency2 <- sort(frequency2, decreasing = TRUE)
words2 <- names(frequency2)

wordcloud(words = words2[1:100], 
          freq = frequency2[1:100], 
          random.order = FALSE, 
          scale = c(5, 0.5),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE, 
          colors = brewer.pal(8, "Dark2"))

frequency2.df <- data.frame(w = factor(words2, 
                                      levels = words2), 
                           f = frequency2)

ggplot(data = frequency2.df[1:20, ], 
       aes(w, f, fill = f),
       xlab = "Terms",
       ylab = "Frequency",
       main = "Most frequent terms") + 
    geom_bar(stat = "identity") + 
    theme(axis.text.x=element_text(angle = 60, hjust = 1))

```

```{r diplayFreq3}
## Trigrams

dtms.sample.trigram.m <- as.matrix(dtms.sample.trigram)

# calculate frequency
frequency3 <- colSums(dtms.sample.trigram.m)

# sort by most frequently used words
frequency3 <- sort(frequency3, decreasing = TRUE)
words3 <- names(frequency3)

wordcloud(words = words3[1:100], 
          freq = frequency3[1:100], 
          random.order = FALSE, 
          scale = c(3, 0.2),
          min.freq = 1,
          rot.per = 0.35, 
          use.r.layout = FALSE, 
          colors = brewer.pal(8, "Dark2"))

frequency3.df <- data.frame(w = factor(words3, 
                                      levels = words3), 
                           f = frequency3)

ggplot(data = frequency3.df[1:20, ], 
       aes(w, f, fill = f),
       xlab = "Terms",
       ylab = "Frequency",
       main = "Most frequent terms") + 
    geom_bar(stat = "identity") + 
    theme(axis.text.x=element_text(angle = 60, hjust = 1))

```

3. How many unique words do you need in a frequency sorted dictionary to cover 50% of all word instances in the language? 90%?  

```{r}
# using dtm with sparse terms
dtm.sample.unigram.m <- as.matrix(dtm.sample.unigram)

# calculate frequency
frequency <- colSums(dtm.sample.unigram.m)

# sort by most frequently used words
frequency <- sort(frequency, decreasing = TRUE)

words.total <- sum(frequency)
words.agg <- 0
words.count <- 1

while(words.agg < 0.5 * words.total) {
    words.agg <- words.agg + frequency[words.count]
    words.count <- words.count + 1
}

words.count50 <- words.count

while(words.agg < 0.9 * words.total) {
    words.agg <- words.agg + frequency[words.count]
    words.count <- words.count + 1
}

words.count90 <- words.count
```

Number of frequent words covering half of the language: **`r words.count50`**  
Number of frequent words covering 90% of the language: **`r words.count90`**  

4. How do you evaluate how many of the words come from foreign languages?  

One of the ways to identify foreign words is by looking up for a word in a language dictionary. This approach is quite straighforward and has disadvantages such as incorrect classification of misspelled words.  
Another possible way is using machine learning algorithms with language profiles. This approach is used in *langid* library.  

5. Can you think of a way to increase the coverage -- identifying words that may not be in the corpora or using a smaller number of words in the dictionary to cover the same number of phrases?  

One of the ways of identifying words that may not be in the corpora can be possible if there is an external dictionary with linguistic markers such as type of the word (noun, verb, adjective, adverb, etc.)  
Using a smaller number of words in the dictionary to cover the same number of phrases can be done by stemming words and suggesting endings based on some algorithm.

## References

https://rstudio-pubs-static.s3.amazonaws.com/31867_8236987cf0a8444e962ccd2aec46d9c3.html

https://deltadna.com/blog/text-mining-in-r-for-term-frequency/

## Appendix
